Maximum Entropy Spectral Analysis: a case study

### 2021-04-14 paper/mesa.tex
intro

	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% ABSTRACT
\begin{abstract}
The Maximum Entropy Spectral Analysis (MESA) method, developed by Burg, provides a powerful tool to perform spectral estimation of a time-series. The method relies on a generalization of Jaynes' maximum entropy principle and provides the means of inferring the spectrum of a stochastic process in terms of the coefficients of some autoregressive process AR(p) of order $p$.
A closed form recursive solution provides an estimate of the autoregressive coefficients as well as of the order $p$ of the process.
We provide a ready-to-use implementation of the algorithm in the form of a python package `memspectrum`. We characterize our implementation by performing a power spectral density analysis on syntethic data (with known power spectral density) and we compare different criteria for stopping the recursion. Furthermore, we compare the performance of our code with the traditional Welch algorithm, using both syntethic data and real gravitational wave data from the LIGO-Virgo collaboration.
We find that, when compared to Welch's method, Burg's method provides a power spectral density (PSD) estimation with a 
systematically lower variance. This is particularly manifest in the case of a little number of data points, making Burg method most suitable to work in this regime.
%Several methods are characterized on synthetic noise having a known power spectral density (PSD) via the computation of the percentage error and then compared.

@@

The study of the properties of stochastic processes is a crucial task in many fields of physics and mathematical physics \sschmidt{Here we should make some examples... Or at least cite something}.
A particularly interesting class of stochastic processes are the so-called \textit{wide-sense} stationary processes. These are stochastic processes that display an invariance of their statistical properties, such as their two-point autocovariance function, with respect to time translation. If $x(t)$ is a wide-sense stationary process, it is completely determined by the knowledge of the autocorrelation function 
\begin{equation}
	C(\tau) = \mathbf{E}[x_t \cdot x_{t+\tau}]
\end{equation}
or equivalently by the knowledge of their \emph{power spectral density} (PSD) $S(f)$. Thanks to the Wiener-Khinchin theorem the two are related by a Fourier transform: 
\begin{equation}\label{eq:psd-autocorrelation}
	S(f) = \int_{-\infty}^{\infty} \textrm{d}\tau C(\tau) e^{-i 2 \pi f \tau}\,.
\end{equation}
In some literature, the PSD is introduced as  
\begin{equation}\label{eq:psd-f-definition}
	\mathbf{E}[\tilde{x}(f) \cdot \tilde{x}(f^\prime)] = S(f) \delta(f-f^\prime)
\end{equation}
without highlighting its connection with the time structure of the process itself, thus masking some important properties that will explore further in what follows. The latter definition in~\eqref{eq:psd-f-definition} gives, however, i) a straightforward interpretation of the PSD: it measures how much signal ``power" is located in each frequency; ii) an operative way of estimating it for an unknown process.  
%Under such assumption, it can be shown [cite something] that the problem of characterizing the noise properties is greatly simplified and its properties are summarized by a single \textit{autocorrelation function}, which describes how an observation at a given time correlates with another observation a later time. More formally, the properties of a zero mean time series $x_t$ are described by the function:
%\begin{equation}
%	R(\tau) = \mathbf{E}[x_t \cdot x_{t+\tau}]
%\end{equation}
%where the average is computed on different times $t$.
%Thus, a stationary noise time series is completely characterized by the autocorellation function.
%\par
%More often, the study is performed in frequency domain: indeed, for stationary noise the components in frequency $\tilde{x}(f)$ of a stationary time series are conditionally independent from each other. It can be shown that:
%\begin{equation}
%	\mathbf{E}[\tilde{x}(f) \cdot \tilde{x}(f^\prime)] = S(f) \delta(f-f^\prime)
%\end{equation}
%This equation defines the function \textit{power spectral density} (PSD) $S(f)$.
%The PSD has a straightforward interpretation: it measures how much signal ``power" is located in each frequency bin. Furthermore, it is not difficult to show that it is the Fourier transform of the autocorrelation function:
%\begin{equation}
%	S(f) = \int_{-\infty}^{\infty} \textrm{d}\tau R(\tau) e^{-i 2 \pi f \tau}
%\end{equation}
%A power spectral density fully describes a stationary noisy time series and this shows how important is in the study of noise.
%\par
%Thus, in many real application, the knowledge of the PSD is crucial for the understanding of the process and indeed algorithms for PSD computation are a standard part of every software package for computational physics.

An ubiquitous method for such computation is due to Welch \cite{Welch1967} and it based on eqs.(\ref{eq:psd-autocorrelation}-\ref{eq:psd-f-definition}).
Basically, the PSD is obtained by slicing the observed realisation $x(t_1),\ldots,x(t_n)$ of the process $x(t)$ into many windowed batches and averaging the squared moduli of their Fourier transforms.
This approach is equivalent \cite{Lomb,Scargle} to taking the Fourier Transform of the windowed sample autocorrelation $\rho_W$, written as
\begin{equation}
    \rho_{W} = \left\{W_0\rho_0,W_{\pm 1}\rho_{\pm 1}, \dots, W_{\pm M}\rho_{\pm M}, 0, 0, \dots \right\},
\end{equation}
where $\rho$ is the empirical autocorrelation and $M$ is the maximum time lag at which the autocorrelation is computed.
The sequence $W$ is a window function that can be chosen in several different way, each choice presenting advantages and disadvantages for the final estimate of the PSD.

The choice of a window function is arbitrary and typically is made by trial and error, until a satisfactory compromise between variance and resolution of the estimate of PSD is reached. A high frequency resolution implies high variance and vice-versa.
Besides the window function, Welch method requires a number of arbitrary choices to be made, as the number of time slices and the overlap between consecutive slices. All the knobs must be tuned by hand and their choice can dramatically affect the PSD estimation.
The final PSD estimation depends on all such choices, hence begging the question of what the ``real" PSD is.

Another problem with this approach is the requirement for the window to be $0$ outside the interval in which the autocorrelation is computed.
We are arbitrarily assuming $\rho_j = 0$ for $j > M$ and modifying the estimate (i.e. the data) if a non-rectangular window is chosen.
Making assumptions on unobserved data and modifying the ones we have at our disposal introduces a piece of ``spurious" information about the process that we are not allowed to use.

An appealing alternative approach, based on the Maximum Entropy principle \cite{JaynesArticle,jaynes2003ptl, Jaynes_MAXENT}, has been put forward by Burg \cite{burg1975maximum}. Being rooted on solid theoretical foundations, we will see that Burg's method, unlike Welch's, does not require any preprocessing of the data and requires very little tuning of the arbitrary parameters, as it provides an iterative closed form expression for the spectrum of a stochastic stationary time series. Furthermore, it embeds the PSD estimation problem into an elegant theoretical framework and makes minimal assumptions on the nature of the data.
Lastly and most importantly, it provides a robust link between spectral density estimation and the field of autoregressive processes. This provide a natural and simple machinery to forecast a time series, thus predicting past observations based on previous ones.

In this paper we discuss the details of the Maximum entropy principle, its application to Power Spectral Density estimation with Burg algorithm and the link between Burg's algorithm and autoregressive process.
Our goal is to bring (again) to public attention Maximum Entropy Spectral analysis, in the hope that it will be widely employed as a way out to the many problem posed by the ubiquitous Welch algorithm (or other similar methods).
To facilitate this goal, we present and describe a new code, \texttt{memspectrum}, that provides a robust python implementation of the algorithm. The code is ready to use and available at the PyPI repository\footnote{
Available at link: \url{https://pypi.org/project/memspectrum/}.
}
We provide a thorough assessment of the performance of our code and we validate our results performing a number of tests. We also compare our results with those of spectral analysis carried out with the standard Welch method.
In order to apply our model on a realistic setting, we analyse some time series of broad interest in the scientific community.

Our paper is organized as follows: blablabla...

\section{Theoretical foundations}
The Maximum Entropy principle (MAXENT) is among the most important results in Bayesian probability theory. It provides a way to uniquely assign probabilities to a phenomenon in a way that best represent our state of knowledge, while at the same time it is non committal with unavailable information. Its domain of application turned out to be wider than expected. In fact, thanks to Burg \cite{burg1975maximum}, this method has also been applied to perform high quality computation of power spectral densities of time series.

@@

\section{Validation of the model}
It is now clear that MESA provides a recursive formula for computing the coefficients $a_k$ in Eq.~(\ref{eq:MESApsd}). The number $M$ of such coefficients is the same as the number of time steps considered in the computation of the autocorrelation $\bar{r}_m$. In a perfect scenario, this would be equal to the number of points the autocorrelation is computed at (equivalent to the length of data considered). However, the computation of high order coefficients of the autocorrelation is unstable and for high enough $n$, the the estimation for  $\bar{n}_m$ shows a very high variance, broadly scaling as $\sim \left(\sqrt{M - m}\right)^{-1}$.

@@

	\begin{acknowledgments}
%We want to thank Maurizio Maugeri and Costanza Rossi for useful discussions.
\sschmidt{Is it fine to ackonowledge people helping with temperature data and seismic data?}

This research has made use of data, software and/or web tools obtained from the Gravitational Wave Open Science Center (\url{https://www.gw-openscience.org/}), a service of LIGO Laboratory, the LIGO Scientific Collaboration and the Virgo Collaboration. LIGO Laboratory and Advanced LIGO are funded by the United States National Science Foundation (NSF) as well as the Science and Technology Facilities Council (STFC) of the United Kingdom, the Max-Planck-Society (MPS), and the State of Niedersachsen/Germany for support of the construction of Advanced LIGO and construction and operation of the GEO600 detector. Additional support for Advanced LIGO was provided by the Australian Research Council. Virgo is funded, through the European Gravitational Observatory (EGO), by the French Centre National de Recherche Scientifique (CNRS), the Italian Istituto Nazionale di Fisica Nucleare (INFN) and the Dutch Nikhef, with contributions by institutions from Belgium, Germany, Greece, Hungary, Ireland, Japan, Monaco, Poland, Portugal, Spain.
	\end{acknowledgments}

>>>>

	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% ABSTRACT
\begin{abstract}
The Maximum Entropy Spectral Analysis (MESA) method, developed by Burg, provides a powerful tool to perform spectral estimation of a time-series. The method relies on a Jaynes' maximum entropy principle and provides the means of inferring the spectrum of a stochastic process in terms of the coefficients of some autoregressive process AR($p$) of order $p$.
A closed form recursive solution provides an estimate of the autoregressive coefficients as well as of the order $p$ of the process.
We provide a ready-to-use implementation of the algorithm in the form of a python package `memspectrum`. We characterize our implementation by performing a power spectral density analysis on syntethic data (with known power spectral density) and we compare different criteria for stopping the recursion. Furthermore, we compare the performance of our code with the ubiquitous Welch algorithm, using both syntethic data and real gravitational wave data from the LIGO-Virgo collaboration.
We find that, when compared to Welch's method, Burg's method provides a power spectral density (PSD) estimation with a 
systematically lower variance. This is particularly manifest in the case of a little number of data points, making Burg's method most suitable to work in this regime.
%Several methods are characterized on synthetic noise having a known power spectral density (PSD) via the computation of the percentage error and then compared.

@@

The study of the properties of stochastic processes is a crucial task in many fields of physics, astronomy, quantitative biology, astronomy as well as engineering. Of particular interest for several areas of physics is the study of stochastic time series. Among those, a special place is occupied by the so-called \textit{wide-sense} stationary processes. These are stochastic processes that display an invariance of their statistical properties, such as their two-point autocovariance function, with respect to time translation. If $x(t)$ is a wide-sense stationary process, it is completely determined by the knowledge of the autocorrelation function 
\begin{equation}
	C(\tau) = \mathbf{E}[x_t \cdot x_{t+\tau}]
\end{equation}
or, equivalently, by the knowledge of their \emph{power spectral density} (PSD) $S(f)$. Thanks to the Wiener-Khinchin theorem the two are related by a Fourier transform: 
\begin{equation}\label{eq:psd-autocorrelation}
	S(f) = \int_{-\infty}^{\infty} \textrm{d}\tau C(\tau) e^{-i 2 \pi f \tau}\,.
\end{equation}
In some literature, especially in the context of gravitational waves physics, e.g. ~\cite{Finn_1992}, the PSD is introduced as  
\begin{equation}\label{eq:psd-f-definition}
	\mathbf{E}[\tilde{x}(f) \cdot \tilde{x}(f^\prime)] = S(f) \delta(f-f^\prime)
\end{equation}
without highlighting its connection with the time structure of the process itself, thus masking some important properties that will explore further in what follows. The latter definition in~\eqref{eq:psd-f-definition} gives, however, i) a straightforward interpretation of the PSD: it measures how much signal ``power" is located in each frequency; ii) an operative way of estimating it for an unknown process.

A ubiquitous method for such computation is due to Welch \cite{Welch1967} and it based on eqs.(\ref{eq:psd-autocorrelation}-\ref{eq:psd-f-definition}).
Basically, the PSD is obtained by slicing the observed realisation $x(t_1),\ldots,x(t_n)$ of the process $x(t)$ into many window-corrected batches and averaging the squared moduli of their Fourier transforms.
This approach is equivalent~\cite{Lomb,Scargle} to taking the Fourier Transform of the windowed sample autocorrelation $\rho_W$, written as
\begin{equation}
    \rho_{W} = \left\{W_0\rho_0,W_{\pm 1}\rho_{\pm 1}, \dots, W_{\pm M}\rho_{\pm M}, 0, 0, \dots \right\},
\end{equation}
where $\rho$ is the empirical autocorrelation and $M$ is the maximum time lag at which the autocorrelation is computed.
The sequence $W$ is a window function that can be chosen in several different way, each choice presenting advantages and disadvantages for the final estimate of the PSD.

The choice of a window function is arbitrary and typically is made by trial and error, until a satisfactory compromise between variance and resolution of the estimate of PSD is reached. A high frequency resolution implies high variance and vice-versa.
Besides the window function, Welch' method requires a number of arbitrary choices to be made, since the number of time slices and the overlap between consecutive slices. All the knobs must be tuned by hand and their choice can dramatically affect the PSD estimation.
The final PSD estimate depends on all such choices, hence begging the question of what the ``real" PSD is.

Another problem with this approach is the requirement for the window to be $0$ outside the interval in which the autocorrelation is computed.
We are arbitrarily assuming $\rho_j = 0$ for $j > M$ and modifying the estimate (i.e. the data) if a non-rectangular window is chosen.
Making assumptions on unobserved data and modifying the ones we have at our disposal introduces ``spurious" information about the process that we in general do not really have.

An appealing alternative approach, based on the Maximum Entropy principle \cite{JaynesArticle,jaynes2003ptl, Jaynes_MAXENT}, has been put forward by Burg \cite{burg1975maximum}. Being rooted on solid theoretical foundations, we will see that Burg's method, unlike Welch's, does not require any preprocessing of the data and requires very little tuning of the algorithm parameters, since it provides an iterative closed form expression for the spectrum of a stochastic stationary time series. Furthermore, it embeds the PSD estimation problem into an elegant theoretical framework and makes minimal assumptions on the nature of the data.
Lastly and most importantly, it provides a robust link between spectral density estimation and the field of autoregressive processes. This provide a natural and simple machinery to forecast a time series, thus predicting future observations based on previous ones.

In this paper, we discuss the details of the Maximum entropy principle, its application to the problem of PSD estimation with Burg algorithm and the link between Burg's algorithm and autoregressive process.
Our goal is to bring (again) to public attention Maximum Entropy Spectral analysis, in the hope that it will be widely employed as a way out to the many problem posed by Welch's algorithm (or other similar methods).
To facilitate this goal, we present and describe a new code, \texttt{memspectrum}, that provides a robust and easy-to-use python implementation of the algorithm\footnote{
Available at link: \url{https://pypi.org/project/memspectrum/}.
}.
We provide a thorough assessment of the performance of our code and we validate our results performing a number of tests on simulated and real data.
We also compare our results with those of spectral analysis carried out with the standard Welch method.
In order to apply our model on a realistic setting, we analyse some time series of broad interest in the scientific community.

Our paper is organized as follows: we being by briefly reviewing the theoretical foundations of the maximum entropy principle in Sec.~\ref{sec:foundations}. Sec.~\ref{sec:validation} presents the validation of Burg's method as well as of our implementation on simulated data. In Sec.~\ref\label{sec:Welch_comparison} we compare the results from \texttt{memspectrum} with the Welch method; Sec.~\ref{sec:applications} presents a few applications to time series and, finally, we conclude with a discussion in Sec.~\ref{sec:conclusion}.

\section{Theoretical foundations}\label{sec:foundations}
The Maximum Entropy principle (MAXENT) is among the most important results in Bayesian probability theory. It provides a way to uniquely assign probabilities to a phenomenon in a way that best represent our state of knowledge, while at the same time it is non committal with unavailable information. Its domain of application turned out to be wider than expected. In fact, thanks to Burg \cite{burg1975maximum}, this method has also been applied to perform high quality computation of power spectral densities of time series.

@@

\section{Validation of the model}\label{sec:validation}
It is now clear that MESA provides a recursive formula for computing the coefficients $a_k$ in Eq.~(\ref{eq:MESApsd}). The number $M$ of such coefficients is the same as the number of time steps considered in the computation of the autocorrelation $\bar{r}_m$. In a perfect scenario, this would be equal to the number of points the autocorrelation is computed at (equivalent to the length of data considered). However, the computation of high order coefficients of the autocorrelation is unstable and for high enough $n$, the the estimation for  $\bar{n}_m$ shows a very high variance, broadly scaling as $\sim \left(\sqrt{M - m}\right)^{-1}$.

@@

	\begin{acknowledgments}
We are grateful to S.~Shore and PLEASE ADD WHOEVER WE INTERACTED WITH IN THIS PAPER for useful discussions. This research has made use of data, software and/or web tools obtained from the Gravitational Wave Open Science Center (\url{https://www.gw-openscience.org/}), a service of LIGO Laboratory, the LIGO Scientific Collaboration and the Virgo Collaboration. LIGO Laboratory and Advanced LIGO are funded by the United States National Science Foundation (NSF) as well as the Science and Technology Facilities Council (STFC) of the United Kingdom, the Max-Planck-Society (MPS), and the State of Niedersachsen/Germany for support of the construction of Advanced LIGO and construction and operation of the GEO600 detector. Additional support for Advanced LIGO was provided by the Australian Research Council. Virgo is funded, through the European Gravitational Observatory (EGO), by the French Centre National de Recherche Scientifique (CNRS), the Italian Istituto Nazionale di Fisica Nucleare (INFN) and the Dutch Nikhef, with contributions by institutions from Belgium, Germany, Greece, Hungary, Ireland, Japan, Monaco, Poland, Portugal, Spain.
	\end{acknowledgments}
